---
title: "Getting started with tidyllm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Your Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Getting Started with tidyllm

tidyllm is an R package designed to provide a unified interface for interacting with various large language model APIs. This vignette will guide you through the basic setup and usage of tidyllm.

### Installation

You can install tidyllm directly from GitHub using devtools:

```{r, eval= FALSE, echo=TRUE}
# Install devtools if not already installed
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

# Install tidyllm from GitHub
devtools::install_github("edubruell/tidyllm")
```

### Setting up API Keys

Before using tidyllm, you need to set up API keys for the services you plan to use. Here's how to set them up for different providers:

1. Anthropic (Claude):
```{r, eval= FALSE, echo=TRUE}
Sys.setenv(ANTHROPIC_API_KEY = "YOUR-ANTHROPIC-API-KEY")
```

2. OpenAI (ChatGPT):
```{r, eval= FALSE, echo=TRUE}
Sys.setenv(OPENAI_API_KEY = "YOUR-OPENAI-API-KEY")
```

3. Groq:
```{r, eval= FALSE, echo=TRUE}
Sys.setenv(GROQ_API_KEY = "YOUR-GROQ-API-KEY")
```

Alternatively, you can set these keys in your `.Renviron` file for persistent storage. For this, execute `usethis::edit_r_environ()`, and add a
line with with an API key in this file, for example:

```{r, eval= FALSE, echo=TRUE}
ANTHROPIC_API_KEY="XX-XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX"
```

If you do not want to use remote models you can also work with local large lange models via `ollama`. Install it from [the official project website](https://ollama.com/). Ollama sets up a local large language model server that you can use to run open-source models on your own devices. 


### Basic Usage

Let's start with a simple example using tidyllm to interact with different language models:

```{r firsttry,  eval=TRUE, echo=TRUE}
library(tidyllm)

# Start a conversation with Claude
conversation <- llm_message("What is the capital of France?") |>
  claude()

#Standard way that llm_messages are printed
conversation

# Get the last reply
last_reply(conversation)

# Continue the conversation with ChatGPT
conversation <- conversation |>
  llm_message("What's a famous landmark in this city?") |>
  chatgpt()

# Get the last reply
last_reply(conversation)
```

### Working with Images

tidyllm also supports sending images to multimodal models:

```{r images,  eval=TRUE, echo=TRUE}
# Describe an image using a llava model on ollama
image_description <- llm_message("Describe this image",
                                 .imagefile = "https://raw.githubusercontent.com/edubruell/tidyllm/refs/heads/main/tidyllm.png") |>
  ollama(.model = "llava")

# Get the last reply
last_reply(image_description)
```

### Sending R Outputs to Language Models

You can include R code outputs in your prompts. In particular you can use `.caputre_plot` to send the last plot pane to a model. 
The `.f`-argument runs a function and sends it output to a model. 

```{r routputs,  eval=TRUE, echo=TRUE}
library(tidyverse)

# Example data
example_data <- tibble(
  x = rnorm(100),
  y = 2 * x + rnorm(100)
)

# Create a plot
ggplot(example_data, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm")

# Send the plot and data summary to a language model
analysis <- llm_message("Analyze this plot and data summary:", 
                        .capture_plot = TRUE,
                        .f = ~{summary(example_data)}) |>
  claude()

last_reply(analysis)
```

### Adding PDFs to messages

The `llm_message()` function also supports extracting text from PDFs and including it in the message. This allows you to easily provide context from a PDF document when interacting with the AI assistant.

To use this feature, you need to have the `pdftools` package installed. If it is not already installed, you can install it with:

```{r,  eval=FALSE, echo=TRUE}
install.packages("pdftools")
```

To include text from a PDF in your prompt, simply pass the file path to the `.pdf` argument of the `chat` function:

```{r,  eval=TRUE, echo=TRUE}
llm_message("Please summarize the key points from the provided PDF document.", 
     .pdf = "https://pdfobject.com/pdf/sample.pdf") |>
     ollama()
```

The package will automatically extract the text from the PDF  and include it in the prompt sent to the an API. The text will be wrapped in `<pdf>` tags to clearly indicate the content from the PDF:

```
Please summarize the key points from the provided PDF document.

<pdf filename="example_document.pdf">
Extracted text from the PDF file...
</pdf>
```

### API parameters

Different API functions support different model parameters like  how deterministic the response should be via parameters like temperature. Please read API-documentation and the documentation of the model functions for specific examples.

```{r temperature,  eval=TRUE, echo=TRUE}
temp_example <- llm_message("Explain how temperature parameters work in large language models.")

#per default it is non-zero
temp_example |> ollama()

#Temperature sets the randomness of the answer
#0 is one extreme where the output becomes fully deterministic. 
#Else the next token is allways sampled from a list of the most likely tokens. Here only the most likely token is used every time.
temp_example |> ollama(.temperature=0) |> last_reply()
temp_example |> ollama(.temperature=0) |> last_reply()# Same answer
```

### Use-case vignettes. 

There are three vignettes that highlight special use-cases of the `tidyllm`-package and 
walk you true a typical use-case for the package:

  1. Structured question answering for pdf documents
  2. Automated image labeling
  3. Synthetic data generation for testing questionnaires
