---
title: "Embedding Models in tidyllm"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE
)
```

While most **tidyllm** use cases revolve around chat models, it also supports **embedding models**, another type of large language model. These models generate numerical vectors that map input text to points in a high-dimensional space, where each point represents the semantic meaning of the text:

- **Similar meanings are close together:** A text about a "cat" and one about a "kitten" map to nearby points.
- **Different meanings are farther apart:** A text about a "cat" and one about a "car" map to points much farther apart.

## Semantic Search in Economics Paper Abstracts

To demonstrate embeddings in action, we implement a **semantic search** on a dataset of **22,960 economics paper abstracts published between 2010 and 2024**. Instead of relying on keyword matching, we use embeddings to find papers with similar topics based on their underlying meaning.

```{r abstracts_file, message = FALSE}
library(tidyverse)
library(tidyllm)
library(here)

abstracts <- read_rds("abstracts_data.rds")
print(abstracts, width = 60)

target_abstract <- "We use the expansion of the high-speed rail network in Germany as a natural experiment to examine the causal effect of reductions in commuting time between regions on the commuting decisions of workers and their choices regarding where to live and where to work. We exploit three key features in this setting: i) investment in high-speed rail has, in some cases dramatically, reduced travel times between regions, ii) several small towns were connected to the high-speed rail network only for political reasons, and iii) high-speed trains have left the transportation of goods unaffected. Combining novel information on train schedules and the opening of high-speed rail stations with panel data on all workers in Germany, we show that a reduction in travel time by one percent raises the number of commuters between regions by 0.25 percent. This effect is mainly driven by workers changing jobs to smaller cities while keeping their place of residence in larger ones. Our findings support the notion that benefits from infrastructure investments accrue in particular to peripheral regions, which gain access to a large pool of qualified workers with a preference for urban life. We find that the introduction of high-speed trains led to a modal shift towards rail transportation in particular on medium distances between 150 and 400 kilometers."
```

With the dataset ready, we use tidyllm to generate embeddings and perform a semantic search to find papers similar to the `target_abstract` from a paper on commuting and high-speed rail by [Heuermann and Schmieder (2018)](https://drive.google.com/file/d/1rKhTsrc9Qsi2gRUQg86zKKuUWvzx5HE1/view). For this task we use the `mxbai-embed-large` model, available via Ollama with `ollama pull mxbai-embed-large`. It is important to choose your embedding model carefully upfront, as each model produces unique numerical representations that are not interchangeable between models.

Embedding APIs are also available for `voyage()`, `mistral()`, `gemini()`, and `openai()` (as well as `azure_openai()`).

### Step 1: Computing Embeddings for One Abstract

To compute an embedding of the target abstract we use `embed()` with `ollama()` as the provider:

```{r, eval = FALSE}
target_tbl <- target_abstract |>
  embed(ollama(.model = "mxbai-embed-large:latest"))

target_tbl
str(target_tbl)
```

```{r, echo = FALSE}
target_tbl <- tibble(
  input      = target_abstract,
  embeddings = list(runif(1024, min = -1, max = 1))
)
print(target_tbl, width = 60)
```

The `embed()` function returns a tibble with two columns:

- **input:** The original text provided for embedding.
- **embeddings:** A list column containing the numerical vector for each input.

In our case we have a single input and the embeddings column contains a 1,024-dimensional vector:

```{r}
str(target_tbl$embeddings)
```

### Step 2: Computing Embeddings for the Entire Abstract Corpus

When working with a large corpus like our 22,960 abstracts, embedding all entries in a single pass is impractical. Commercial APIs typically cap inputs per request at 50 to 100; local APIs face memory and compute constraints.

We batch the data into groups of 200 and write results to disk as we go. On a MacBook Pro M1 Pro this completes in approximately 25 minutes and writes 207 MB of data. To compare multiple target abstracts against the collection, we only need to embed the corpus once.

```{r, eval = FALSE}
generate_abstract_embeddings <- function(abstracts) {
  embedding_batches <- tibble(abstract = abstracts) |>
    group_by(batch = floor(1:n() / 200) + 1) |>
    group_split()

  n_batches <- length(embedding_batches)
  glue("Processing {n_batches} batches of 200 abstracts") |> cat("\n")

  embedding_batches |>
    walk(\(batch) {
      batch_number <- pull(batch, batch) |> unique()
      glue("Generate Text Embeddings for Abstract Batch: {batch_number}/{n_batches}") |> cat("\n")

      batch$abstract |>
        embed(ollama(.model = "mxbai-embed-large:latest")) |>
        write_rds(here("embedded_abstracts", paste0(batch_number, ".rds")))
    })
}

abstracts |>
  pull(abstract) |>
  generate_abstract_embeddings()
```

```{r, echo = FALSE}
"Processing 115 batches of 200 abstracts
Generate Text Embeddings for Abstract Batch: 1/115
Generate Text Embeddings for Abstract Batch: 2/115
Generate Text Embeddings for Abstract Batch: 3/115
...\n" |> cat()
```

After the function finishes, load the precomputed embeddings:

```{r, eval = FALSE}
embedded_abstracts <- here("embedded_abstracts") |>
  dir() |>
  map_dfr(\(f) read_rds(here("embedded_abstracts", f)))
```

### Step 3: Performing the Semantic Search

With embeddings precomputed, we perform a semantic search using cosine similarity. **Cosine similarity** measures the similarity between two vectors by computing the cosine of the angle between them; it ranges from **-1** (opposite directions) to **1** (identical directions):

\[
\text{cosine\_similarity}(\mathbf{a}, \mathbf{b}) = \frac{\sum_{i=1}^n a_i b_i}{\sqrt{\sum_{i=1}^n a_i^2} \cdot \sqrt{\sum_{i=1}^n b_i^2}}
\]

Vectors at 90 degrees (orthogonal, no semantic overlap) have cosine similarity of 0; vectors pointing in the same direction have similarity close to 1. We express this as a simple function:

```{r}
cosine_similarity <- function(a, b) {
  sum(a * b) / (sqrt(sum(a^2)) * sqrt(sum(b^2)))
}
```

We apply it to find the 10 most similar abstracts in the corpus:

```{r, eval = FALSE}
top10_similar <- embedded_abstracts |>
  mutate(cosine_sim = map_dbl(embeddings,
                              \(e) cosine_similarity(e, target_tbl$embeddings[[1]]))) |>
  arrange(desc(cosine_sim)) |>
  slice(1:10) |>
  rename(abstract = input) |>
  left_join(abstracts |> select(year, authors, journal, abstract), by = "abstract") |>
  select(-embeddings)
```

The ten most similar articles in the corpus:

```{r echo = FALSE}
read_rds("top10_asbtracts.rds") |>
  mutate(abstract = stringr::str_wrap(abstract, width = 60)) |>
  kableExtra::kable("html", caption = "Top 10 Most Similar Abstracts") |>
  kableExtra::kable_styling(full_width = FALSE,
                            bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

Unsurprisingly, the top result is the target abstract itself. The remaining results strongly align with the target's focus on transportation infrastructure and its economic effects; the second result discusses German high-speed rail, while the third and fourth cover Japan's Shinkansen and China's HSR network. This illustrates the model's ability to capture semantic connections across diverse geographic and policy contexts.

## Multimodal Embeddings

While most embedding functions work only on text, the Voyage AI provider in tidyllm supports embedding text and images in the same space. This is particularly useful for **cross-modal search**, where text descriptions and images need to be compared semantically.

Use `img()` to create image objects and mix them with text in a list. When passing such a list to `embed(voyage())`, the function automatically routes to Voyage's multimodal API:

```{r, eval = FALSE, echo = TRUE}
list("a dish consisting of a sausage served in the slit of a partially sliced bun",
     img(here("hotdog.jpg"))) |>
  embed(voyage())
```

```{r, echo = FALSE, eval = TRUE}
tibble::tibble(
  input = c(
    "a dish consisting of a sausage served in the slit of a partially sliced bun",
    "[IMG] hotdog.jpg"
  ),
  embeddings = purrr::map(1:2, \(i) runif(min = -1, max = 1, n = 1024))
)
```

Since the text description and the image refer to the same concept, their embeddings will be close together in the high-dimensional space, enabling similarity-based retrieval across modalities.

## Reranking with Voyage

Embedding-based search is fast for large corpora but not always precise; cosine similarity scores can be noisy when the query and documents differ in style or length. **Reranking** complements the retrieval step: given a short candidate list, a reranker scores each document against the query with a more powerful cross-attention model.

`voyage_rerank()` wraps Voyage's `/v1/rerank` endpoint and returns a tibble sorted by relevance score:

```{r, eval = FALSE}
candidates <- top10_similar$abstract

voyage_rerank(
  .query     = target_abstract,
  .documents = candidates,
  .model     = "rerank-2",
  .top_k     = 5
)
```

```{r, echo = FALSE}
tibble::tribble(
  ~index, ~relevance_score, ~document,
  0L,     0.921,            "We study the economic effects of German High-Speed Rail...",
  4L,     0.887,            "This paper examines how Shinkansen expansion affected...",
  1L,     0.854,            "We estimate the effect of China's high-speed rail network...",
  7L,     0.731,            "Infrastructure investment and regional labor markets...",
  2L,     0.698,            "Commuting zones and the spatial structure of local labor..."
)
```

A typical workflow combines both: embed the full corpus once, retrieve the top 50 candidates with cosine similarity, then rerank to surface the best 5 to 10. This keeps compute costs low while improving final result quality.

## Outlook: Clustering and Beyond

Embeddings open the door to a wide range of analytical techniques beyond semantic search:

- **Clustering for topic discovery:** Methods like K-means or hierarchical clustering can automatically group texts into themes, useful for exploratory research on large corpora without predefined labels.
- **Dimensionality reduction and visualisation:** PCA, t-SNE, or UMAP project embeddings into 2D or 3D, revealing clusters, trends, and outliers at a glance.
- **Retrieval-augmented generation (RAG):** An embedding search retrieves semantically similar documents, which are then appended to the model's prompt to ground responses in relevant evidence.
- **Supporting qualitative research workflows:** As discussed in [Kugler et al. (2023)](https://www.iaw.edu/iaw-diskussionspapiere.html?file=files/dokumente/ab%20Januar%202023/iaw_dp_143.pdf), embeddings can automate coding steps in qualitative research, though models can still struggle with implicit references and closely related themes that human researchers handle naturally.
