---
title: "Structured Question Answering from PDFs"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE
)
```

Navigating through a large collection of academic papers can be time-consuming, especially when you're trying to extract specific insights or determine the relevance of each document to your research. With **tidyllm**, you can streamline this process by automating the extraction of structured answers directly from PDF documents using large language models.

Imagine you have a folder of papers on the economic effects of generative AI, and you need to assess how each paper is related to your own research interests. This article guides you through setting up a workflow that processes the first few pages of each paper, asks an AI model targeted questions, and returns the answers in a structured format, perfect for converting into a table for easy review and analysis.

## Example Workflow

Imagine your folder looks something like this, many downloaded papers but no structure yet:

```{r setup, message = FALSE}
library(tidyverse)
library(tidyllm)
dir("aipapers")
```

Our goal is to get a first overview of these papers and to give them good file names.

### Step 1: Generating Messages for Each Document

First, we prepare a list of messages for all PDFs in the folder by applying `llm_message()` with a prompt to the first five pages of each document. Even though `gpt-5.4` can process up to 128,000 tokens (roughly 80-90 pages of English text), we limit the input to five pages for demonstration purposes; the introduction is usually enough to get a first overview of a paper.

Attach PDFs with `pdf_file()` and the `.media` argument. On providers that support native PDF rendering (Claude, Gemini), the file is sent as binary, preserving images, tables, and formatting. On all other providers the text is extracted automatically and wrapped in `<pdf>` tags in the prompt. Pass `.text_extract = TRUE` to always force text extraction regardless of the provider:

```{r document_tasks, eval = FALSE}
files <- list.files("aipapers", full.names = TRUE, pattern = "\\.pdf$")

document_tasks <- files |>
  map(\(f) llm_message(
    "Below are the first 5 pages of a document.
     Summarise the document based on the provided schema.",
    .media = pdf_file(f, pages = 1:5)
  ))
```

### Step 2: Defining the Schema for Structured Output

We define a schema that outlines the expected data types for each field in the model's responses. This enables the model to return answers in a consistent, structured format that converts directly into a tibble.

```{r document_schema}
document_schema <- tidyllm_schema(
  name             = "DocumentAnalysisSchema",
  Title            = field_chr("The full title of the provided document"),
  Authors          = field_chr("A semicolon-separated list of authors"),
  SuggestedFilename = field_chr("Suggest a filename in the format \"ReleaseYear_Author_etal_ShortTitle.pdf\". Use the publication year if available; otherwise, use XXXX."),
  Type             = field_fct("Is the document a Policy Report or a Research Paper based on its style, structure, and purpose?",
                               .levels = c("Policy", "Research")),
  Empirics         = field_chr("A 100 word description of empirical methods used, including data collection, statistical techniques, or analysis methods mentioned."),
  Theory           = field_chr("A 100 word outline of the primary theoretical framework or models discussed, if any."),
  MainPoint        = field_chr("A one-sentence summary of the main point raised"),
  Contribution     = field_chr("A short explanation of the main contributions claimed in the document"),
  KeyCitations     = field_chr("The four most frequently cited or critical references in the first 5 pages")
)
```

`field_chr()` returns text, `field_dbl()` returns numbers, and `field_fct()` restricts to allowed categories via the `.levels` argument. Setting `.vector = TRUE` on any field returns a list of values; here we keep a flat structure so each document maps to a single-row tibble. `name` is a special field providing an identifier for the schema.

### Step 3: Running the Analysis on a Sample Document

To test the setup, we run the analysis on a single document with `chat()` and the schema:

```{r example_task, eval = FALSE}
example_task <- document_tasks[[1]] |>
  chat(openai(.json_schema = document_schema,
              .model       = "gpt-5.4"))
```

```{r example_task_hidden, echo = FALSE}
example_task <- read_rds("example_task.rds")
```

### Step 4: Extracting and Formatting the Results

`get_reply_data()` extracts the model's structured response:

```{r example_task_get_reply}
get_reply_data(example_task)
```

The model returns a named list matching our schema, which `as_tibble()` converts to a single row. We can also inspect token usage:

```{r example_task_get_metadata, eval = FALSE}
get_metadata(example_task)
```

```{r, echo = FALSE}
tibble(
  model             = "gpt-5.4",
  timestamp         = as.POSIXct("2026-03-16 10:00:00"),
  prompt_tokens     = 4265L,
  completion_tokens = 337L,
  total_tokens      = 4602L,
  stream            = FALSE
) |>
  mutate(api_specific = list(list(
    system_fingerprint        = "",
    completion_tokens_details = list(),
    prompt_tokens_details     = list()
  )))
```

At current `gpt-5.4` batch pricing the per-document cost is a few cents; processing a folder of 50 papers costs well under a dollar.

### Step 5: Scaling Up to a Whole Batch of Papers

After confirming the single-document analysis works, we extend the workflow to process a full batch. Batch APIs from Anthropic and OpenAI offer up to 50% savings compared to single requests, and batch rate limits are separate from standard per-model limits so they don't affect your regular quota.

```{r batch, eval = FALSE}
document_tasks |>
  send_batch(openai(.json_schema = document_schema,
                    .model       = "gpt-5.4")) |>
  write_rds("document_batch.rds")
```

The output of `send_batch()` contains the input message list plus batch metadata (a `batch_id` attribute and unique per-message identifiers). Batches are processed within 24 hours; the example here completed in about 10 minutes.

> ⚠️ **Note:** Save the RDS file to disk. It acts as a checkpoint, letting you reload and check batch status or retrieve results across R sessions without resubmitting the batch.

Check whether a batch has completed:

```{r, eval = FALSE}
read_rds("document_batch.rds") |>
  check_job(openai())
```

You can also list all OpenAI batches with `list_batches(openai())` or view them in the [OpenAI batches dashboard](https://platform.openai.com/batches/).

### Step 6: Getting Data from the Entire Batch

Once the batch is complete, fetch all responses with `fetch_batch()`:

```{r, eval = FALSE}
results <- read_rds("document_batch.rds") |>
  fetch_batch(openai())
```

```{r results_hidden, echo = FALSE}
results <- read_rds("document_results.rds")
```

Map `get_reply_data()` and `as_tibble()` over the batch output to produce a tidy table:

```{r}
document_table <- results |>
  map(get_reply_data) |>
  map_dfr(as_tibble)

document_table
```

This table can be exported to Excel with `writexl::write_xlsx()` for further review, or used to programmatically rename the PDFs using the model's suggested filenames.

## Further Notes on Working with Large Documents

### Context Length

When working with long documents, context length limits how much text the model can process in a single query. If a document exceeds this limit, the model only sees a portion of it and may miss important sections.

Focusing on the first five pages, as in this workflow, typically captures the abstract, introduction, and methodology; enough for a first overview. For documents where later sections matter, split them into smaller chunks and process each separately. The overlap between chunks helps preserve continuity.

### Gemini and Claude for Image-Heavy PDFs

When you use `pdf_file()` with `gemini()` or `claude()`, the PDF is sent as binary rather than extracted text. This means the model sees the actual layout, tables, charts, and scanned pages, not just the raw characters. For text-only workflows on any provider, pass `.text_extract = TRUE` to `pdf_file()` to force text extraction and skip binary encoding entirely.

For large or frequently reused documents you can also upload them once to the provider's servers with `upload_file()` and reference the returned handle in messages via `.files`. This avoids re-sending the binary on every request. Provider file APIs often accept a broader range of formats beyond PDFs: Gemini supports video, audio, and various document types; OpenAI accepts Office files (DOCX, PPTX, XLSX), source code, ZIP archives, and more. This makes `upload_file()` useful for workflows that go beyond plain PDFs.

```{r, eval = FALSE}
report <- upload_file(claude(), .path = "annual_report.pdf")

llm_message("Summarise the key risks.", .files = report) |>
  chat(claude())
```

See the [Working with Files and Media](https://edubruell.github.io/tidyllm/articles/tidyllm_video.html) article for a full walkthrough of the file upload workflow across providers.

### Local Models

Local models via `ollama()` or `llamacpp()` are a strong choice for privacy-sensitive documents, as no data leaves your machine. Modern models like `qwen3.5` support a context window of up to 262,144 tokens in principle; the practical limit on a typical laptop is RAM. On a 16 GB machine a 9B model at Q4 quantisation already uses roughly 6 GB for weights, leaving limited headroom for long contexts before inference slows to a crawl or the system starts swapping.

For most academic papers the first five pages fit comfortably within what laptop hardware can handle. For longer documents, limit the page range or reduce the context window explicitly via the `.num_ctx` parameter in `ollama()`.

> ⚠️ **Note:** When using paid remote APIs, consider data privacy, especially for sensitive documents. Local models via `ollama()` or `llamacpp()` keep all processing on-device.

## Outlook

This structured question-answering workflow streamlines extraction of key insights from academic papers and adapts directly to other document-heavy tasks; reports, policy documents, contracts, or news archives can all be processed with the same pattern of schema definition, single-document testing, and batch scale-up.
