---
title: "Classifying Texts with tidyllm"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE#,
#  comment = "#>"
)
```
> ⚠️ **Note:**  This vignette refers to the development version of **tidyllm** 0.3.0. which is not yet on CRAN

Classification tasks are a key challenge when dealing with unstructured data in surveys, customer feedback, or for image analysis. This article introduces a practical, step-by-step guide to using **tidyllm** for classifying text, streamlining the process.

## A Common Classification Task

Imagine you’ve just collected thousands of survey responses where people describe their jobs in their own words. Some responses are detailed, others are vague, and there’s plenty of variation in between. Now, you need to categorize these into standardized occupation codes, like those from the SOC classification system of the Bureau of Labor Statistics. Manually sorting through each response could take days, if not weeks, and inconsistencies between coders are almost guaranteed. For instance, your dataset might look something like this: 7,000 rows of occupation descriptions, ranging from "Librarian" to "Making sure everything runs"

```{r firstchunk, message = FALSE}
library(tidyllm)
library(tidyverse)
library(glue)
occ_data <- read_rds("occupation_data.rds")
occ_data
```
Our goal is to classify these messy responses into one of these 22 2-digit occupation codes:
```{r occ_code_tibble, message = FALSE}
occ_codes <- read_rds("occ_codes_2digits.rds") |>
  print(n=Inf)
```

In this article, we take a structured approach to tackle the classification task efficiently. Here's a step-by-step workflow for this kind of task:

1. **Classifying and Manually Correcting a Sub-Sample**:
   - **Pick a Sample**:Start by filtering the dataset to retain only distinct occupation descriptions. Then, randomly select a sample of distinct responses to work with. 
   - **Initial Classification**: Use a simple prompt to categorize these responses into occupation codes.
   - **Manual Correction**: Review and correct the classifications to create a reliable ground truth.

2. **Optimizing the Classification Process**:
   - **Training/Test Split**: Again split this ground truth dataset using `rsample` into training and test sets.
   - **Experimentation**: Test different prompts, models, and parameters on the training set, comparing one-shot and multi-shot approaches.
   - **Model Evaluation**: Use `yardstick` to find the best-performing combination on the training data.
   - **Testing**: Apply the best-performing model to the test set to evaluate how well it performs on unseen occupation descriptions.

3. **Scaling Up to the Full Dataset**:
   - **Full Classification**: Use the validated model setup to classify the entire dataset efficiently.

## Classifying a Sub-Sample

We start by ensuring we only classify distinct responses. This eliminates duplicates and ensures a more efficient and reliable classification process:
```{r distinct_occ}
# Pick only distinct occupations from the dataset
distinct_occupations <- occ_data |> 
  distinct(occupation = occupation_open)

print(distinct_occupations, n = 5)
```
This will help us focus on variations across distinct occupations, avoiding repeated classification efforts on identical responses. 

Next, we divide the distinct occupations into a sub-sample for manual classification and a remaining portion to be used later. We use the `initial_split()` function from the `rsample` package, splitting 10% of the data into a smaller test set for manual correction and model training:
```{r split_subsample}
#Set a seed for reproducability
set.seed(123)

# Split the distinct occupations into a sub-sample (10%) and the rest (90%)
library(rsample)
occ_split <- initial_split(distinct_occupations, prop = 0.8)

# Retrieve the sub-sample and the remaining data
rest_of_data <- training(occ_split)
sub_sample <- testing(occ_split)

print(sub_sample, n = 5) 
```
By splitting the data, we now have a smaller sub-sample of 422 observations to work with during the initial classification stage. 

### A simple sequential classifier

To classify this sub-sample of occupation descriptions, we first generate prompts with a specific classification task for each occupation. Each prompt directs the model to classify the occupation based on a numerical code, ensuring that the output aligns with our coding system. We use `glue()` to create prompts for each occupation description to the prompt with `{occupation}` and `purrr::map()` to pass each prompt to `llm_message()`. The output is a list of messages ready for submission to a large language model.

```{r classify_occ_prompts}
prompts <- glue('
      Classify this occupation response from a survey: {occupation}
      
      Pick one of the following numerical codes from this list. 
      Respond only with the code!
      11 = Management Occupations                                    
      13 = Business and Financial Operations Occupations             
      15 = Computer and Mathematical Occupations                     
      17 = Architecture and Engineering Occupations                  
      19 = Life, Physical, and Social Science Occupations            
      21 = Community and Social Service Occupations                  
      23 = Legal Occupations                                         
      25 = Educational Instruction and Library Occupations           
      27 = Arts, Design, Entertainment, Sports, and Media Occupations
      29 = Healthcare Practitioners and Technical Occupations        
      31 = Healthcare Support Occupations                            
      33 = Protective Service Occupations                            
      35 = Food Preparation and Serving Related Occupations          
      37 = Building and Grounds Cleaning and Maintenance Occupations 
      39 = Personal Care and Service Occupations                     
      41 = Sales and Related Occupations                             
      43 = Office and Administrative Support Occupations             
      45 = Farming, Fishing, and Forestry Occupations                
      47 = Construction and Extraction Occupations                   
      49 = Installation, Maintenance, and Repair Occupations         
      51 = Production Occupations                                    
      53 = Transportation and Material Moving Occupations
      99 = Missing Occupation (No clear occupation)', occupation = sub_sample$occupation)

classification_tasks <- map(prompts,llm_message)
```
A typical message from this list of classification tasks looks like this: 
```{r classification_tasks1}
classification_tasks[[1]]
```
To classify an initial subset, which will need to be manually verified later, it is beneficial to use a reliable commercial model, as its output typically requires fewer adjustments. Here, we choose Claude-3.5-Sonnet, setting `.temperature` to 0 to ensure deterministic (non-random) responses. Initially, we test our prompt with the standard sequential `chat(claude())` approach. This preliminary step allows us to verify that the model interprets the prompt accurately and produces output in the required format before running the full batch of classification tasks.

To this end, we create a tibble containing the occupation descriptions from our sub-sample alongside the prepared classification tasks. We then randomly select 10 rows from this tibble. Each selected occupation is passed sequentially through the `classify_sequential()` function that passes each classification task through `chat(claude())`,
fetches the reply with `get_reply()` and converts the text output of the model to a numeric code with `parse_number()`. Finally, the results are joined with the `occ_codes` table to map the numeric occupation codes to their corresponding occupation labels, providing an interpretable  output:

```{r classification_tasks1_output, eval=FALSE}
classify_sequential <- function(occupation_open,message){
    raw_code <- message |>
        chat(claude(.temperature = 0)) |>
        get_reply() |>
        parse_number()
    
    tibble(occupation_open=occupation_open, occ2=raw_code)
}

tibble(occupation_open = sub_sample$occupation, message = classification_tasks) %>%
  slice_sample(n=10) |>
  pmap_dfr(classify_sequential) |>
  left_join(occ_codes, by="occ2")
```
```{r out_t1, echo=FALSE}
tribble(
  ~occupation_open,              ~occ2, ~occ_title,
  "Fine Carpentry",               47,   "Construction and Extraction Occupations",
  "Eyeglass makin'",              51,   "Production Occupations",
  "Layin' down shingles",         47,   "Construction and Extraction Occupations",
  "Handle construction budgets",  13,   "Business and Financial Operations Occupations",
  "Bodyshop guy",                 49,   "Installation, Maintenance, and Repair Occupations",
  "Hair dresser",                 39,   "Personal Care and Service Occupations",
  "Sort and deliver mail",        43,   "Office and Administrative Support Occupations",
  "Ops oversight",                11,   "Management Occupations",
  "Oversee all cleaning operations", 11, "Management Occupations",
  "Systems administrator",        15,   "Computer and Mathematical Occupations"
)
```
This output demonstrates that the model accurately interprets the prompts, assigning each occupation description to a suitable category. Another check before we  classify the entire set of messages in `classification_tasks` using Anthropic's Claude-3.5-Sonnet model, is to look at the token-usage and the cost of a typical message with `get_metdadata()`:

```{r, eval=FALSE}
classification_tasks[[1]] |>
  chat(claude(.temperature = 0)) |>
  get_metadata()
```
```{r, eval=TRUE,echo=FALSE}
tibble(model="claude-3-5-sonnet-20241022",
       timestamp=NA,
       prompt_tokens = 349,
       completion_tokens = 5,
       total_tokens = 354)
```
Assuming all messages in this initial batch are roughly similar, we can expect all 442 messages to need around 150,000 prompt tokens and 2200 completion tokens. At the time of writing, the price for one million input tokens on Claude was 3\$, while one million output tokens did cost 15\$. Prices using batch request are only half. Classifying our subsample with batch-processing costs roughly 22 cents.

### Using a claude batch request

Batch-processing offers a significant cost advantage over sequential classification. Anthropic charges half the price per token for batch requests compared to single requests, making it an efficient choice for large-scale tasks. While batch requests do not return results instantly, they are typically processed within 24 hours, often much faster. A Message Batch is limited to either 10,000 requests or 32 MB in size, whichever is reached first. 

To initiate a batch request, we use the `send_claude_batch()` function (you can of course also use the more general verb `send_batch(claude())`. This batch submission function returns the same list of classification tasks that we uploaded, but marked with an attribute that contains a batch-id from the Claude API as well as unique names for each list element that can be used to stitch together messages with replies, once the batch is ready.  If you supply a named list of messages, these names are used as identifiers (ensuring they are unique before submission). For convenient access later, we save the batch output locally, allowing us to resume processing from the batch even after closing the R session:
```{r batch_send, eval=FALSE}
classification_tasks |>
  send_claude_batch(.temperature = 0) |>
  write_rds("sub_sample_batch.rds")
```

We can check the status of a batch with:
```{r check_batch, eval=FALSE}
read_rds("sub_sample_batch.rds") |>
  check_claude_batch()
```

```{r check_results, eval =TRUE, echo=FALSE}
tibble::tribble(
  ~batch_id,                          ~status, ~created_at,           ~expires_at,           ~req_succeeded, ~req_errored, ~req_expired, ~req_canceled,
  "msgbatch_015JWDHzEPP8k7qzWNC9vCY3", "ended", as.POSIXct("2024-10-31 16:33:35"), as.POSIXct("2024-11-01 16:33:35"), 442, 0, 0, 0
)
```
We now see that our batch was created and its status is `ended`. All 442 requests succeeded. Batch results can be downloaded within  29 days after creation. You can see all available batches in the API with `list_claude_batches()` or in the Anthropic console. To download the results of our batch we use the `fetch_claude_batch()` function:

```{r fetch_batch, eval=FALSE}
occ2_codes <- read_rds("sub_sample_batch.rds")) |>
  fetch_claude_batch() |>
  map_chr(get_reply) |>
  parse_number()

tibble(occupation_open = sub_sample$occupation,
       occ2 = occ2_codes) |>
  left_join(occ_codes, by="occ2") |>
  writexl::write_xlsx("ground_truth_excel.xlsx")
```
We get a list of messages with assistant replies from `fetch_claude_batch()` that we pass into `map_chr(get_reply)` to get assistant replies from each message as a character vector. We then again parse the numeric code and merge occupation titles. To ensure that the output is correct, we can export the results to Excel with `write_xlsx` from the `writexl` package and manually fix miss-classifications. 

A manual review of the classifier’s output showed highly promising results. Out of 443 classifications, only 9 needed corrections, indicating an error rate of just 2% for the `claude()`-based classifier with our initial prompt. Most issues arose from not correctly identifying unclear responses as missing, such as "Doin' the numbers" or "Barster" (a potential mix of Barrister or Barkeeper). 

At this point we could stop and proceed to classify the entire data with the `send_claude_batch()`, given its strong performance and manual validation on unseen data and that scaling up to the data would cost less than 1.50$. However, to illustrate the general principle how to find good LLM-based classifiers when you are less sure, we will now use the initial ground truth we have built as a base to experiment with different models and prompts. For example, we could try to determine if simpler alternatives or small local models would perform just as well. 

## Optimizing and Testing Classifiers
Next, we present a general approach to optimize and test different LLM-based classifiers. First, we split the dataset into training and test sets using `rsample` to ensure that we can experiment with different prompts and setups on training data but only evaluate our final model performance on unseen data.

```{r, eval=TRUE, echo=TRUE}
ground_truth <- readxl:::read_xlsx("ground_truth_corrected.xlsx")

# Split the ground-truth into training and testing sets
set.seed(123)
gt_split <- initial_split(ground_truth, prop = 0.7)

# Retrieve training and testing data
train_data <- training(gt_split)
test_data  <- testing(gt_split)

print(train_data,n=5)
```
###  Writing a flexible classifier function
To test different prompts and models systematically we need to create a flexible classifier function that can handle different prompts or models that can work with any api-function including local models that do not support batch requests.  This function can take the open `occupation` question, our ground-truth for `occ2` a `prompt` and `prompt_id`, a `provider` and a `model` as arguements allowing us to test each occupation classification across a grid of prompts and models.

```{r classify_occ2, eval=FALSE, echo=TRUE}
# External numerical code list for reusability
numerical_code_list <- c('
      11 = Management Occupations                                    
      13 = Business and Financial Operations Occupations             
      15 = Computer and Mathematical Occupations                     
      ... abreviated ...
      51 = Production Occupations                                    
      53 = Transportation and Material Moving Occupations
      99 = Missing Occupation')

# Classification function that accepts prompt, api_function, and model 
# as well as the ground truth to pass through as arguments
classify_occupation_grid <- function(occupation,
                                occ2,
                                prompt,
                                prompt_id,
                                provider,
                                model){
  # Output what the model is currently doing to the console
  glue("Classifying: {model} - {prompt_id} - {occupation}\n") |> cat("\n")
  
  # List of valid codes as strings
  valid_codes <- as.character(occ_codes$occ2)
  
  # Initialize classification_raw
  classification <- tryCatch({
    # Get the assistant's reply using the dynamically provided API function and model
    assistant_reply <- llm_message(prompt) |>
      chat(provider(.model = model, .temperature = 0)) |>
      get_reply() |>
      str_squish()
    
    # Validate the assistant's reply
    if (assistant_reply %in% valid_codes) {
      as.integer(assistant_reply)
    } else {
      98L  # Return 98 for invalid responses
    }
  }, error = function(e){
    97L  # Return 97 in case of an error (e.g., API failure)
  })
  
  # Return a tibble containing the original occupation description and classification result
  tibble(
    occupation_open = occupation,
    occ2_predict    = classification,
    occ2_truth      = occ2,
    model           = model,
    prompt_id       = prompt_id
  )
}
```
```{r numlist, eval=TRUE, echo=FALSE}
numerical_code_list <- c('
      11 = Management Occupations                                    
      13 = Business and Financial Operations Occupations             
      15 = Computer and Mathematical Occupations                     
      17 = Architecture and Engineering Occupations                  
      19 = Life, Physical, and Social Science Occupations            
      21 = Community and Social Service Occupations                  
      23 = Legal Occupations                                         
      25 = Educational Instruction and Library Occupations           
      27 = Arts, Design, Entertainment, Sports, and Media Occupations
      29 = Healthcare Practitioners and Technical Occupations        
      31 = Healthcare Support Occupations                            
      33 = Protective Service Occupations                            
      35 = Food Preparation and Serving Related Occupations          
      37 = Building and Grounds Cleaning and Maintenance Occupations 
      39 = Personal Care and Service Occupations                     
      41 = Sales and Related Occupations                             
      43 = Office and Administrative Support Occupations             
      45 = Farming, Fishing, and Forestry Occupations                
      47 = Construction and Extraction Occupations                   
      49 = Installation, Maintenance, and Repair Occupations         
      51 = Production Occupations                                    
      53 = Transportation and Material Moving Occupations
      99 = Missing Occupation')
```

### Defining the Prompt and Model Grid

We define a set of prompts and models that we want to test. This will allow us to apply the classifier across different configurations and compare results. Here's how the prompts and models are set up:

1. **Prompts:**
  - **Prompt 1:** A detailed prompt, asking the model to classify occupations and warning it not to make guesses.
  - **Prompt 2:** Explicitly ask to handle invalid responses by returning a special code (99) when the input does not resemble a valid occupation.
  - **Prompt 3:** A shorter, more concise version to test whether the model performs similarly with less detailed instructions.
2. **Models:**
  - **Llama3.2:3B:** A opensource large language model with just 3 billion parameters that is very fast, when you run it locally via `ollama()`
  - **Gemma2:9B:** Another candidate model, which performs well for classification tasks, but is more than double the size of Lama3.2 and is therefore somewhat slower.

We set up a grid combining all the prompts and models. The `expand_grid()` function from the tidyverse is a useful tool here to create every possible combination of prompts and models, which we will use to evaluate the classifier:

```{r gridsetup, eval=TRUE,echo=TRUE}
prompts <- tibble(prompt = 
         c( #Original prompt
            'Classify this occupation response from a survey: {occupation}
            Pick one of the following numerical codes from this list. 
            Respond only with the code!
            {numerical_code_list}',
            #Explicit instruction to avoid classifying something wrong
           'Classify this occupation response from a survey: {occupation}
            Pick one of the following numerical codes from this list. 
            Respond only with the code!
            {numerical_code_list}
           
            If this does not look like a valid occupation response reply with 
            just 99
           ',
           #Shorter prompt
           'Classify this occupation: {occupation}. 
           Respond only with one of the following codes: 
           {numerical_code_list}'
         ),
         prompt_id = 1:3)


grid <- expand_grid(train_data, 
                    prompts, 
                    model = c("llama3.2", "gemma2")) |>
  arrange(model) %>% # Arrange by model so ollama does not reload them often
  rename(occupation = occupation_open) |>
  rowwise() |>  # Glue together prompts and occupation row-by-row
  mutate(prompt = glue(prompt)) |>
  ungroup() |> # Ungroup after the rowwise operation
  select(model,occupation,occ2,prompt_id,prompt)

nrow(grid)
```
To run the classification across the entire grid, we use `pmap_dfr()` from the purrr package, which allows us to iterate over multiple arguments simultaneously. Each combination of occupation response, prompt, and model is passed into the `classify_occupation()` function, and the results are concatenated into a single tibble:
```{r gridrun, eval=FALSE,echo=TRUE}
grid_results <- grid |>
  pmap_dfr(classify_occupation_grid,
           provider = ollama)
```
```{r, eval=TRUE, echo=FALSE}
classifications <- c(
  "gemma2 - 1 - Computer network technician", 
  "gemma2 - 2 - Computer network technician", 
  "gemma2 - 3 - Computer network technician", 
  "..."
)

cat(glue::glue("Classifying: {classifications}"), sep = "\n")
```
> ⚠️ **Note:** Running an extensive classification grid like this, especially with large datasets or slow models, can take a significant amount of time. Therefore, it’s often reasonable to save intermediate results periodically, so that you don’t lose progress if something goes wrong (e.g., a crash or a network issue). By combining `pwalk()` with `save_rds()`, you can run each combination of the grid independently and store results incrementally. 

After we run our grid we get some insights into how well models and prompts work on the train data. Here we can experiment with different models and parameters as much as we want and see what works. 

### Accuracy estimates

To create an overview of prediction accuracy overview we use the `yardstick` package.
For some functions in  `yardstick`, we also need to encode the ground truth and the model predictions as factors:
```{r grid_results, eval=TRUE,echo=FALSE}
grid_results <- bind_rows(
  read_rds(here::here("vignettes","articles","gemma_results.rds")),
  read_rds(here::here("vignettes","articles","llama32_results.rds"))
)
```
```{r grfactors, eval=TRUE,echo=TRUE}
library(yardstick)

gr_factors <- grid_results |>
  mutate(across(starts_with("occ2_"),
                ~factor(.x,
                        levels=c(occ_codes$occ2,97, 98,99),
                        labels=c(occ_codes$occ_title,"APIFAIL","INVALID","MISSING")
                ))) 
```

First we just just calculate and plot the model accuracy:
```{r accuracy, fig.width=16, fig.height=9, fig.alt="Bar plot showing accuracy by prompt ID and model."}
accuracy <- gr_factors  |>
  group_by(prompt_id, model) |>
  accuracy(truth = occ2_truth, estimate = occ2_predict)

accuracy |>
  ggplot(aes(x = as.factor(prompt_id), y = .estimate, fill = model)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Accuracy by Prompt and Model", x = "Prompt ID", y = "Accuracy", fill="") +
  theme_bw(22) +
  scale_y_continuous(labels = scales::label_percent(),limits = c(0,1)) +
  scale_fill_brewer(palette = "Set1")
```

Here are the main insights of our classification experiment: 

- **Gemma2** consistently outperforms **llama3.2**, but even **gemma2's** top performance on *Prompt 1 (79.1)* falls far short of Claude Sonnet's **98% accuracy** in the initial run.
- The simplified prompt clearly introduces challenges, especially for the smaller models like **llama3.2**, which shows particularly poor performance (as low as 20.2% on Prompt 3). This suggests that these models might not generalize well to slightly varied or less explicit prompts, whereas Claude is able to handle the variations with far greater ease.

Let's look at the confusion matrix for the top **gemma2** performance to see whether any specific occupation category causes problems here:
```{r confmatrix, fig.width=16, fig.height=12, fig.alt="A plot of a confusion matrix for the occupation classification in the main tested model."}
conf_mat <- gr_factors |>
  filter(prompt_id == 1, model == "gemma2") |>
  conf_mat(truth = occ2_truth, estimate = occ2_predict)

# Autoplot the confusion matrix
autoplot(conf_mat, type = "heatmap") +
  scale_fill_gradient(low = "white", high = "#E41A1C") +
  ggtitle("Confusion Matrix for Model gemma2, Prompt 1") +
  theme_bw(22) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
A lot of the miss-classifications are in management occupations and office services, missing occupations are again not classified well, but classified into other occupations. So there is nothing here that seems like an immediate easy fix. Yet, there are other tricks to try that go beyond simple one-shot classification. A common idea is to use [multi-shot classifications](https://docs.anthropic.com/en/docs/build-with-claude/prompt-engineering/multishot-prompting#example-analyzing-customer-feedback) (and give the model classification examples in the prompt, or even guide its output by letting it complete conversations where it answers the right way). The function `df_llm_message()` let's you easily built such message histories where you put some words into the mouth of the model. But there are other ways that can help you increase accuracy. A popular one is to first let the model reason in detail what the best candidates for classification would be to build a chain of thought before it gives the final answer. 

### Multistep Chain-of-Thought Prompting
For this we need a major change in the classification function, because we send two messages to the model. A first one elicits a reasoning step, the second one asks for the final code, based on the answer to the first message. This is easily doable with the message-chaining abilities of **tidyllm**

Here is a modified chain of thought prompting function with a reasoning and an output step:

```{r cot, eval=FALSE, echo=TRUE}
classify_occupation_cot <- function(occupation,
                                    occ2,
                                    provider,
                                    model,
                                    stream = FALSE){
  # Output what the model is currently doing to the console
  glue("Classifying with CoT: {model} - {occupation}\n") |> cat("\n")
  
  # Step 1: Ask the model to think through the problem
  prompt_reasoning <- glue('
    Think about which of the following occupation codes would best describe this occupation description from a survey respondent: "{occupation}"
    
    {numerical_code_list}
    
    Explain your reasoning for the 3 top candidate codes step by step. Then evaluate which seems best.
  ')
  
  reasoning_response <- tryCatch({
    conversation <<- llm_message(prompt_reasoning) |>
      chat(provider(.model = model, .temperature = 0, .stream=stream))
    
    conversation |>
      get_reply()
  }, error = function(e){
    conversation <<- llm_message("Please classify this occupation response: {occupation}")
    "Error in reasoning step."
  })
  
  # Step 2: Ask the model to provide the final answer
  prompt_final <- glue('
    Based on your reasoning, which code do you pick? Answer only with a numerical code!

  ')
  
  final_response <- tryCatch({
    conversation |>
    llm_message(prompt_final) |>
      chat(provider(.model = model, .temperature = 0, .stream=stream)) |>
      get_reply() |>
      str_squish()
  }, error = function(e){
    "97"
  })
  
  # Validate the model's final response
  valid_codes <- as.character(occ_codes$occ2)
  
  classification <- if (final_response %in% valid_codes) {
    as.integer(final_response)
  } else {
    98L  # Return 98 for invalid responses
  }
  
  # Return a tibble containing the original occupation description and classification result
  tibble(
    occupation_open = occupation,
    occ2_predict    = classification,
    occ2_truth      = occ2,
    model           = glue("{model}_cot"),
    reasoning       = reasoning_response,
    final_response  = final_response
  )
}
```
Note that this increases the compute time for **gemma2** massively because it now produces a multiple of the output from before. Each run produces roughly a page of reasoning, before the second question generates a number, while our original prompting strategy just gave us one number. So run-times increase strongly.

> **Note:** To test more complicated prompting strategies like this in realtime, another feature of **tidyllm** comes in handy. You can stream back responses  just like in online chatbot interfaces during a first test run with the `.stream`-argument for most provider functions in `chat()` to immediately see whether you made strange errors in your prompting. Therefore, here we pass the stream arguement trough `classify_occupation_cot()` to first test the function and the prompting on single inputs before we scale it up to the whole data. 

Let's run this function  with **gemma2**:

```{r, eval=FALSE, echo=TRUE}
results_cot <- grid |>
  filter(model=="gemma2", prompt_id ==1) |>
  select(-prompt,-prompt_id) |>
  pmap_dfr(classify_occupation_cot,api_function = ollama, stream=FALSE)
```
```{r, eval=TRUE, echo=FALSE}
classifications <- c(
  "gemma2 - 1 - Computer network technician", 
  "gemma2 - 2 - Educational support", 
  "gemma2 - 3 - Fine Carpentry", 
  "..."
)

cat(glue::glue("Classifying with CoT: {classifications}"), sep = "\n")
```

Did this work? It turns out in this case it did not!
```{r cot_hidden, eval=TRUE,echo=FALSE}
results_cot <- read_rds("gemma_cot_results.rds")
```
```{r cot_acc, eval=TRUE,echo=TRUE}
results_cot |>
  mutate(across(starts_with("occ2_"),
                ~factor(.x,
                        levels=c(occ_codes$occ2,97, 98,99),
                        labels=c(occ_codes$occ_title,"APIFAIL","INVALID","MISSING")
                )))  |>
  accuracy(truth = occ2_truth, estimate = occ2_predict)
```
The accuracy is even worse than without the reasoning step! One interesting thing though is that the model has the right classification in one of its candidates in the reasoning step for 94% of the cases. So probably, there is still some way to get a better final result out of **gemma2**:

```{r cot_acc2, eval=TRUE,echo=TRUE}
results_cot |>
  rowwise() |>
  mutate(reasoning_classifications = str_extract_all(reasoning,"\\d{2}"),
         reasoning_classifications = list(map_int(reasoning_classifications, as.integer)),
         right_in_reasoning = occ2_truth %in% reasoning_classifications
         ) |>
  ungroup() |>
  count(right_in_reasoning) |>
  mutate(pct=n/sum(n))
```

However, in our specific case we already  had a clear favorite final choice without even needing to do further testing. We already know the 98% test performance for Claude Sonnet 3.5, since we manually checked it. So no further check of prompt or models on the test data is needed. We can just run our entire sample through `claude_send_batch()`

## Conclusion

In this article, we've demonstrated how `tidyllm` can be effectively used to tackle complex text classification tasks. By walking through a typical process step-by-step, from sample selection and prompt design to model testing and optimization, we've highlighted the flexibility of the package in dealing with real-world data. 

