---
title: "Classifying Texts with tidyllm"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
options(width = 90)
```

You have thousands of short free-text answers and a fixed list of categories. A language model can sort them in minutes, but the result is less solid than it looks. The model decides what a label means from your instructions, and a vague instruction gets a different reading from case to case. Most current frontier models also no longer let you set `temperature`, so asking the same question twice does not guarantee the same answer. And a label that the model gives every time can still be wrong.

This article builds a workflow that treats these as things to measure: a codebook the model can follow, labels the model cannot misspell, repeated runs to see where it disagrees with itself, a second pass for the disagreements, and a human-labelled sample to check the whole thing against. The central point is that agreement is not accuracy. Repeated runs tell you how stable the model is; only a comparison with human labels tells you whether it is right.

Every call that needs an API key is shown in a chunk that is not evaluated. Right after it, a second chunk loads the saved results of that call and computes the numbers and tables, so you can follow the article without a key and see what the real run produced.

## The task

Seven thousand survey respondents described their job in their own words. We want each answer in one of the 22 two-digit SOC occupation groups, or in code 99 when the answer names no occupation.

```{r setup, message = FALSE}
library(tidyllm)
library(tidyverse)
library(glue)

occ_data  <- read_rds("occupation_data.rds")
occ_codes <- read_rds("occ_codes_2digits.rds")

distinct_occupations <- occ_data |>
  distinct(occupation = occupation_open) |>
  mutate(id = row_number())

print(distinct_occupations, n = 5)
```

Classifying each distinct answer once is enough; the labels are joined back to the respondents at the end.

## Why a single answer is not enough

Earlier versions of this article set `.temperature = 0` and treated the reply as fixed, which no longer holds. Claude 5 models, recent Gemini models and several OpenAI models reject or ignore sampling parameters, and `chat(openai(), .temperature = 0)` stops with an error for them. `provider_capabilities()` shows which provider functions still accept the argument. That does not mean the model uses it: Gemini 3.6 Flash and later, including the default model of `gemini()`, accept `.temperature` and ignore it.

Even where `temperature` is accepted, the literature on LLM annotation reports that repeated runs of the same prompt on the same text can return different labels, and that the instability is largest on borderline texts (Barrie, Palaiologou and Törnberg, 2024; Haldar and Hockenmaier, 2025). The practical answer in applied work is the one social scientists already use for human coders: label more than once, measure agreement, and look closely at the disagreements. Separately, an LLM label is only as good as its agreement with humans on your task, so the workflow always ends with a validation sample (Pangakis, Wolken and Fasching, 2023).

We use `claude(.model = "claude-sonnet-5-5")` throughout and set no temperature.

## Step 1: Write a codebook, not a list

A bare list of codes leaves the model to guess what "Management" or "Missing" means. A codebook gives each label a definition, a decision rule for overlaps, and one or two examples. A human coder would need the same thing.

```{r codebook}
codebook <- read_file("classifier_codebook.txt") |> trimws()
cat(codebook)
```

Keep the codebook in a file under version control. When you change it, you have changed the measurement, and the validation in step 6 has to be run again. The last section of this article shows what happens when you edit the codebook to fix the errors you found.

## Step 2: Constrain the answer with a schema

Asking for "only the code" and parsing the reply with `parse_number()` breaks when the model adds a word. A schema removes the problem: `field_fct()` makes the label an enum, so the provider only accepts one of the listed codes. We add two more fields. A short rationale gives you something to read when you audit a row, and a three-level confidence gives a coarse signal to sort by. Do not treat the self-reported confidence as a probability; step 4 shows how much it is worth.

```{r schema}
label_levels <- as.character(c(occ_codes$occ2, 99)) |> unique()

occ_schema <- tidyllm_schema(
  name       = "occupation_label",
  rationale  = field_chr("One sentence: which words in the answer decide the group."),
  occ2       = field_fct("The SOC two-digit group from the codebook.", .levels = label_levels),
  confidence = field_fct("How clear the answer is.", .levels = c("low", "medium", "high"))
)
```

```{r prompt_fn}
make_prompt <- function(occupation) {
  llm_message(glue(
    'You code survey answers into occupation groups.
     Use only the codebook below. Choose the single best group.
     If the answer names no occupation, use 99.

     Codebook:
     {codebook}

     Survey answer: "{occupation}"'
  ))
}
```

Look at the request before spending anything on it. `.dry_run = TRUE` returns the request object that would be sent.

```{r dry_run, eval = FALSE}
make_prompt(distinct_occupations$occupation[1]) |>
  chat(claude(.model = "claude-sonnet-5-5", .json_schema = occ_schema, .dry_run = TRUE))
```

Enum-constrained output guarantees a valid label, not a correct one. Some work reports that heavy output constraints can cost accuracy on hard cases (Tam et al., 2024, on format restrictions and reasoning), which is why the rationale field exists and why step 6 measures accuracy.

One call looks like this:

```{r one_call, eval = FALSE}
make_prompt("Doin' the numbers") |>
  chat(claude(.model = "claude-sonnet-5-5", .json_schema = occ_schema)) |>
  get_reply_data()
```

`get_reply_data()` returns a plain named list with one element per schema field. The label stays a character string, because "99" and "11" are labels and not numbers.

## Step 3: Label every answer several times

Because replies can differ between runs, we ask for each answer `k` times and keep every run. Three runs is the smallest number that gives a majority; five gives a finer agreement score at a higher price. We validate on a random sample of 150 distinct answers, so the full dataset is never sent while we are still learning how well the method works. `parallel_chat()` runs the requests concurrently and returns the replies in input order, with the names of the input list kept, so we name each task `id__run` and split the name afterwards.

```{r repeat_tasks, eval = FALSE}
k <- 3

sample_df <- distinct_occupations |>
  slice_sample(n = 150)

tasks <- sample_df |>
  cross_join(tibble(run = seq_len(k))) |>
  mutate(task_id = paste(id, run, sep = "__"))

llms <- tasks$occupation |>
  map(make_prompt) |>
  set_names(tasks$task_id)
```

```{r repeat_run, eval = FALSE}
replies <- parallel_chat(
  llms,
  claude(.model = "claude-sonnet-5-5", .json_schema = occ_schema),
  .max_active = 4
)
write_rds(replies, "runs_claude.rds")
```

Failed requests do not stop the call. `parallel_chat()` warns how many failed, and the failed elements of the list hold the error condition instead of a message. The extraction below turns those into missing values, so no row disappears silently.

```{r tidy_runs, eval = FALSE}
extract_run <- function(reply) {
  if (!inherits(reply, "tidyllm::LLMMessage")) {
    return(tibble(rationale = NA_character_, occ2 = NA_character_, confidence = NA_character_))
  }
  as_tibble(get_reply_data(reply))
}

runs <- map(replies, extract_run) |>
  bind_rows(.id = "task_id") |>
  separate(task_id, c("id", "run"), sep = "__", convert = TRUE) |>
  rename(label = occ2)
```

The saved results of this run hold three labels for each of the 150 sampled answers.

```{r load_runs}
sample_df <- read_rds("classifier_sample.rds")
runs      <- read_rds("classifier_runs.rds")
run_meta  <- read_rds("classifier_runs_meta.rds")

runs |>
  left_join(sample_df, by = "id") |>
  select(id, occupation, run, label, confidence) |>
  print(n = 6)
```

The token counts of the 450 calls come from the response metadata of each call. The codebook is part of every request, so the prompt is much longer than the answer it labels.

```{r tokens}
tibble(
  calls                = nrow(run_meta$tokens),
  prompt_tokens        = sum(run_meta$tokens$input_tokens),
  completion_tokens    = sum(run_meta$tokens$output_tokens),
  prompt_per_call      = round(mean(run_meta$tokens$input_tokens)),
  completion_per_call  = round(mean(run_meta$tokens$output_tokens))
)
```

### The same job as a batch

For a full dataset where a result an hour later is acceptable, the same list goes through `send_batch()` at half the price. `check_batch()` reports the status, and `fetch_batch()` returns the replies once the batch has ended. The names of the list survive the round trip, so the extraction code is nearly unchanged. We sent the first 20 requests of the sample as a batch to test it.

```{r batch_variant, eval = FALSE}
job <- send_batch(llms[1:20], claude(.model = "claude-sonnet-5-5", .json_schema = occ_schema))
write_rds(job, "runs_batch_job.rds")

job <- read_rds("runs_batch_job.rds")
check_batch(job, claude())

replies <- fetch_batch(job, claude())

runs <- map(replies, get_reply_data) |>
  bind_rows(.id = "task_id") |>
  separate(task_id, c("id", "run"), sep = "__", convert = TRUE) |>
  left_join(distinct_occupations, by = "id")
```

Here are the two `check_batch()` outputs of that batch, the first right after submission and the second three minutes later:

```
# A tibble: 1 x 8
  batch_id          status      created_at          expires_at          req_succeeded
  <chr>             <chr>       <dttm>              <dttm>                      <int>
1 msgbatch_017iBx1… in_progress 2026-10-02 13:22:40 2026-10-03 13:22:40             0
# i 3 more variables: req_errored <int>, req_expired <int>, req_canceled <int>

# A tibble: 1 x 8
  batch_id          status created_at          expires_at          req_succeeded
  <chr>             <chr>  <dttm>              <dttm>                      <int>
1 msgbatch_017iBx1… ended  2026-10-02 13:22:40 2026-10-03 13:22:40            20
# i 3 more variables: req_errored <int>, req_expired <int>, req_canceled <int>
```

The batch reported 27,823 prompt tokens for its 20 requests, about 1,390 per request. That matches the live calls above, because the codebook dominates the prompt. The saved batch results can be compared with the first live run of the same answers:

```{r batch_results}
batch_runs <- read_rds("classifier_batch_results.rds")

batch_runs |>
  select(id, occupation, label = occ2, confidence) |>
  inner_join(
    runs |> filter(run == 1) |> select(id, live_label = label),
    by = "id"
  ) |>
  summarise(answers = n(), same_label_as_live_run = sum(label == live_label))
```

### Using more than one model

Repeating one model measures how stable it is with itself. A different model measures something else: whether two independent systems read the codebook the same way. Run the same `llms` through a second provider and add its labels as extra runs. Agreement across models is a stronger signal than agreement within one, because the failures of two unrelated models are less correlated (Ziems et al., 2024, report model-dependent results across tasks). We did not run this here.

```{r second_model, eval = FALSE}
replies_gemini <- parallel_chat(llms, gemini(.json_schema = occ_schema), .max_active = 4)
```

## Step 4: Summarise agreement and flag disagreements

For each answer we take the majority label, the share of runs that voted for it, and a flag for rows where the runs did not all agree.

```{r vote}
vote <- runs |>
  filter(!is.na(label)) |>
  count(id, label, name = "votes") |>
  group_by(id) |>
  arrange(desc(votes), .by_group = TRUE) |>
  summarise(
    n_runs     = sum(votes),
    majority   = first(label),
    agreement  = first(votes) / sum(votes),
    candidates = paste(label, collapse = "/"),
    .groups    = "drop"
  ) |>
  left_join(sample_df, by = "id") |>
  mutate(unstable = agreement < 1)

vote |>
  count(unstable) |>
  mutate(share = n / sum(n))
```

Almost every answer gets the same label in all three runs. Three runs of the same prompt barely change the label, so on this task the repetition buys little by itself. Its value is the short list of rows where the runs differ, and step 6 shows that this list catches only part of the errors.

Two points about reading this table. A tie, such as one vote each for three labels, has no majority; `first()` picks one arbitrarily, so treat every row with `agreement < 0.5` as unresolved. And a unanimous label can still be wrong: agreement shows where the model is unsure of itself and says nothing about the codebook.

The overall agreement across runs can be summarised with Krippendorff's alpha, which handles missing values and works for any number of runs. Here the raters are the `k` runs and the units are the answers. The function below implements alpha for nominal labels in base R, so no extra package is needed.

```{r alpha}
kripp_nominal <- function(m) {
  m  <- m[, colSums(!is.na(m)) >= 2, drop = FALSE]
  lv <- sort(unique(as.vector(m)))
  o  <- matrix(0, length(lv), length(lv))
  for (j in seq_len(ncol(m))) {
    v   <- m[, j]
    v   <- v[!is.na(v)]
    tab <- table(factor(v, lv))
    o   <- o + (tab %o% tab - diag(tab)) / (length(v) - 1)
  }
  n <- sum(o)
  1 - (n - 1) * (n - sum(diag(o))) / (n^2 - sum(rowSums(o)^2))
}

runs |>
  select(run, id, label) |>
  pivot_wider(names_from = id, values_from = label) |>
  select(-run) |>
  as.matrix() |>
  kripp_nominal() |>
  round(3)
```

Alpha near 1 means the runs agree beyond chance. Common rules of thumb call 0.8 reliable and 0.667 the lowest value worth drawing tentative conclusions from (Krippendorff, 2004), but the right bar depends on what you will do with the labels. A high alpha here says the model is consistent with itself. It says nothing yet about the humans.

### Self-reported confidence

The confidence field is cheap, so it is worth checking whether it carries any signal. We can only do that once we have human labels, so here we load the maintainer's hand-checked labels for the 150 sampled answers. They were produced AI-assisted and then checked by hand; step 6 explains how they were made.

```{r load_gold}
gold <- read_rds("classifier_gold.rds")

scored <- vote |>
  left_join(gold, by = "id") |>
  mutate(match = majority == gold)

runs |>
  left_join(select(scored, id, match), by = "id") |>
  group_by(majority_matches_gold = match) |>
  summarise(runs = n(), share_low_confidence = mean(confidence == "low"))
```

Runs on rows where the majority label disagrees with the human label are much more often marked "low". That makes the field useful for ranking rows to read. It does not make it a probability: many mismatched rows still carry "medium" or "high".

### Token probabilities, where the provider offers them

For models that return token log probabilities, the probability of the label token is a continuous confidence score that costs nothing extra. In tidyllm, `openai()` and the Chat Completions based providers accept `.logprobs` and `.top_logprobs`; `claude()` has no such argument, so check `provider_capabilities()` for your provider first. Log probabilities describe one run of one model, so they complement agreement across runs and do not replace it. They are also less informative with structured output, where the label token sits inside a JSON string.

## Step 5: Reclassify the disagreements

The rows that need a second look are the ones where the runs disagreed or where any run said "low". That is a small, concentrated set, which makes a second pass affordable. Instead of voting again with the same prompt, change the task: show the model the answer, the labels the first runs produced and their rationales, and ask for a decision with a stronger setting. Here we use a more capable model with reasoning turned on.

```{r reclassify_prompt}
make_review_prompt <- function(occupation, rationales) {
  llm_message(glue(
    'Earlier independent attempts coded this survey answer; they disagreed or were unsure.
     Decide using the codebook. If the answer is genuinely too vague to place, use 99.

     Codebook:
     {codebook}

     Survey answer: "{occupation}"

     Earlier labels and reasons:
     {rationales}'
  ))
}
```

```{r reclassify_run, eval = FALSE}
any_low <- runs |>
  group_by(id) |>
  summarise(any_low = any(confidence == "low"))

to_review <- vote |>
  left_join(any_low, by = "id") |>
  filter(unstable | any_low) |>
  left_join(
    runs |>
      group_by(id) |>
      summarise(rationales = paste0("- ", label, ": ", rationale, collapse = "\n")),
    by = "id"
  )

review_llms <- map2(to_review$occupation, to_review$rationales, make_review_prompt) |>
  set_names(to_review$id)

review_replies <- parallel_chat(
  review_llms,
  claude(
    .model      = "claude-opus-5-5",
    .json_schema = occ_schema,
    .thinking   = TRUE,
    .max_tokens = 4000
  ),
  .max_active = 4
)

reviewed <- map(review_replies, extract_run) |>
  bind_rows(.id = "id") |>
  transmute(
    id                = as.integer(id),
    review_label      = occ2,
    review_rationale  = rationale,
    review_confidence = confidence
  )
```

Keep `.max_tokens` generous when reasoning is on; the thinking tokens count against it.

```{r reclassify_load}
reviewed <- read_rds("classifier_reviewed.rds")

any_low <- runs |>
  group_by(id) |>
  summarise(any_low = any(confidence == "low"))

to_review <- vote |>
  left_join(any_low, by = "id") |>
  filter(unstable | any_low)

tibble(
  flagged_rows       = nrow(to_review),
  runs_disagreed     = sum(to_review$unstable),
  only_low_confidence = sum(!to_review$unstable),
  reviewed_rows      = nrow(reviewed),
  same_ids           = setequal(to_review$id, reviewed$id)
)
```

```{r final_labels}
final <- vote |>
  left_join(reviewed, by = "id") |>
  left_join(gold, by = "id") |>
  mutate(
    occ2   = coalesce(review_label, majority),
    source = case_when(
      !is.na(review_label) & review_confidence == "low" ~ "human_queue",
      !is.na(review_label)                              ~ "reviewed",
      TRUE                                              ~ "unanimous"
    )
  )

final |> count(source)
```

The review changed the label of a few rows:

```{r review_changes}
final |>
  filter(!is.na(review_label), review_label != majority) |>
  select(occupation, majority, review_label, gold, review_confidence)

tibble(
  changed       = sum(final$review_label != final$majority, na.rm = TRUE),
  reviewed_rows = sum(!is.na(final$review_label))
)
```

The review is a second opinion from the same family of models, so it can repeat a shared misreading. Rows that the reviewer itself marks as low confidence go to a human queue; that is a short list to read by hand.

```{r human_queue}
final |>
  filter(source == "human_queue") |>
  select(occupation, candidates, review_label, gold)
```

## Step 6: Validate against human labels

None of the above tells you whether the labels are right. Only a comparison with human coding does. Draw a random sample of answers, have one or two people code them from the same codebook without seeing the model labels, and resolve their differences. Two rules keep the check honest. The sample must be held out: do not take the codebook examples from it, and do not tune the codebook on it and then report the same sample as your accuracy. And draw it at random across all answers, not only from the flagged ones, or the estimate will be too pessimistic.

For this article the 150 sampled answers were coded in a first pass with AI assistance and then checked by hand by the maintainer. Code 98 in the original file, used for answers that name no occupation, was recoded to 99 to match the codebook. Comparing the model with these labels turned up four labels where the model was right and the first hand label was wrong; they were corrected before the numbers below. That is a useful result in itself: a validation sample is a measurement too, and it has errors that the comparison can surface.

Accuracy and Cohen's kappa come next, computed in base R. Kappa corrects for chance agreement, which matters with 22 groups of very different sizes, where accuracy is dominated by the frequent classes.

```{r metrics}
cohen_kappa <- function(truth, estimate) {
  lv <- sort(unique(c(truth, estimate)))
  tb <- table(factor(truth, lv), factor(estimate, lv))
  po <- sum(diag(tb)) / sum(tb)
  pe <- sum(rowSums(tb) * colSums(tb)) / sum(tb)^2
  (po - pe) / (1 - pe)
}

first_run <- runs |>
  filter(run == 1) |>
  select(id, first_run = label) |>
  inner_join(gold, by = "id")

bind_rows(
  tibble(
    pipeline = "first run only",
    accuracy = mean(first_run$first_run == first_run$gold),
    kappa    = cohen_kappa(first_run$gold, first_run$first_run)
  ),
  tibble(
    pipeline = "majority of 3 runs",
    accuracy = mean(final$majority == final$gold),
    kappa    = cohen_kappa(final$gold, final$majority)
  ),
  tibble(
    pipeline = "majority plus review",
    accuracy = mean(final$occ2 == final$gold),
    kappa    = cohen_kappa(final$gold, final$occ2)
  )
) |>
  mutate(across(where(is.double), \(x) round(x, 3)))
```

The three rows are close. Voting over three runs and the review pass move the accuracy only slightly, because the model gave the same label in nearly all runs to begin with. The review only touches the flagged rows, and those are the hardest answers, so the gain there starts from a low base:

```{r review_effect}
final |>
  filter(!is.na(review_label)) |>
  summarise(rows = n(), before_review = mean(majority == gold), after_review = mean(occ2 == gold))
```

About one answer in six does not match the human label. Part of the errors sit in the flagged rows and part in rows where the model was unanimous:

```{r errors_by_source}
final |>
  group_by(source) |>
  summarise(answers = n(), mismatches = sum(occ2 != gold), accuracy = mean(occ2 == gold))
```

Where do the mismatches come from?

```{r mismatches}
final |>
  filter(occ2 != gold) |>
  arrange(gold) |>
  select(occupation, model = occ2, human = gold, candidates) |>
  print(n = 30)
```

Two groups dominate. The first is vague answers such as "Keep things organized", "Workforce Optimization" or "Buildin and fixin things", where the model chose 99 because the codebook says to prefer 99 over a guess, while the human coder placed the answer in a group from context. The second is the opposite: slang or terse answers such as "Private dick", "I haul" and "Barster", where the model guessed a group and the human coded 99. Between them sit real codebook ambiguities: whether bakery work is production (51) or food service (35), where "Lead mechanic supervisor" belongs, and whether massage is healthcare or personal care.

None of these is random noise that more runs would average away. Most of the mismatches are rows where the model gave the same label every time, which is why the repetition in step 3 cannot find them. They are a mix of questions about what the codebook should say and answers that no coder could place with confidence.

When you compare a model with human coders, compare it with the agreement between two humans on the same sample. If two careful coders only agree 0.85 with each other, a model at 0.85 is not failing; the categories are fuzzy. The model is one more coder whose reliability you estimate, and the evidence so far is that its quality depends on the task, the codebook and the data, so it has to be checked each time (Gilardi, Alizadeh and Kubli, 2023; Pangakis, Wolken and Fasching, 2023). Here only one human checked the labels, so that comparison is not possible, and the 83% should be read with that limit in mind.

Check also whether errors are spread evenly. If one group, one kind of respondent or one phrasing carries most of them, a downstream estimate that uses these labels can be biased even when overall accuracy looks high. Where the labels feed a regression or a share estimate, the validation sample can also be used to correct for misclassification instead of only reporting it.

```{r errors_by_group}
final |>
  group_by(human_group = gold) |>
  summarise(answers = n(), accuracy = mean(occ2 == gold)) |>
  arrange(accuracy, desc(answers)) |>
  slice_head(n = 8)
```

## Step 7: Do not tune the codebook on your validation sample

The mismatch table invites an obvious next step: edit the codebook so that these answers come out right. We tried it. Based on the mismatches above, version 2 of the codebook adds a rule for bakery and food batch work (51 unless it is retail counter work), puts massage under personal care (39), says that lead and supervisor titles and general operations work count as management (11), and counts electrical and mechanical upkeep as installation and repair (49). The 150 answers were labelled again three times with the new codebook.

```{r codebook_v2_run, eval = FALSE}
codebook_v2 <- read_file("classifier_codebook_v2.txt") |> trimws()

make_prompt_v2 <- function(occupation) {
  llm_message(glue(
    'You code survey answers into occupation groups.
     Use only the codebook below. Choose the single best group.
     If the answer names no occupation, use 99.

     Codebook:
     {codebook_v2}

     Survey answer: "{occupation}"'
  ))
}

replies_v2 <- tasks$occupation |>
  map(make_prompt_v2) |>
  set_names(tasks$task_id) |>
  parallel_chat(
    claude(.model = "claude-sonnet-5-5", .json_schema = occ_schema),
    .max_active = 4
  )
```

On the sample the edit was written from, version 2 looks better:

```{r codebook_v2_sample}
majority_label <- function(x) names(which.max(table(x)))

runs_v2 <- read_rds("classifier_runs_v2.rds")

sample_comparison <- bind_rows(
  runs    |> group_by(id) |> summarise(label = majority_label(label)) |> mutate(codebook = "v1"),
  runs_v2 |> group_by(id) |> summarise(label = majority_label(label)) |> mutate(codebook = "v2")
) |>
  inner_join(gold, by = "id") |>
  group_by(codebook) |>
  summarise(answers = n(), accuracy = mean(label == gold))

sample_comparison
```

That gain is not evidence that the new codebook is better, because we wrote the edit while looking at these very errors. To find out, we took 292 further distinct answers with human labels that were not in the sample and labelled each once with each codebook.

```{r codebook_heldout_run, eval = FALSE}
heldout_tasks <- heldout |>
  cross_join(tibble(codebook = c("v1", "v2"))) |>
  mutate(task_id = paste(id, codebook, sep = "__"))
```

```{r codebook_heldout}
heldout_runs <- read_rds("classifier_heldout_runs.rds")
heldout_gold <- read_rds("classifier_heldout_gold.rds") |>
  mutate(gold = ifelse(gold == "98", "99", gold))

heldout_comparison <- heldout_runs |>
  inner_join(select(heldout_gold, id, gold), by = "id") |>
  group_by(codebook) |>
  summarise(answers = n(), accuracy = mean(label == gold))

heldout_comparison
```

On the held-out answers version 2 is no better than version 1; it is slightly worse. The edits fitted the quirks of the 150 answers we looked at. This is the same problem as overfitting a model to a validation set: the sample you tune on stops being a measurement. If you want to improve the codebook, split the human-labelled answers first, tune on one part and report on the other, or label a fresh sample after every round of edits.

## Cost and size

The token table in step 3 gives the numbers for one prompt. A request with this codebook uses roughly 1,400 prompt tokens and about 75 completion tokens, depending on how long the rationale runs. At that size the full dataset would look like this:

```{r scale}
per_call <- run_meta$tokens |>
  summarise(prompt = mean(input_tokens), completion = mean(output_tokens))

price <- c(input = 2, output = 10)

tibble(
  distinct_answers  = nrow(distinct_occupations),
  runs              = 3,
  requests          = distinct_answers * runs,
  prompt_tokens     = round(requests * per_call$prompt),
  completion_tokens = round(requests * per_call$completion),
  usd_standard      = round((prompt_tokens * price[["input"]] + completion_tokens * price[["output"]]) / 1e6, 2),
  usd_batch         = usd_standard / 2
)
```

The prices are those of `claude-sonnet-5-5` on 2026-10-02: 2 USD per million input tokens and 10 USD per million output tokens, half of that through the batch API ([Anthropic pricing](https://platform.claude.com/docs/en/about-claude/pricing)). Prices change, so replace `price` with your provider's current rates. At these rates the 450 calls behind this article's three runs cost about 1.60 USD, the second pass with `claude-opus-5-5` (4 and 20 USD) about 0.20 USD, and the codebook experiment in step 7 about 3.90 USD. Check your own numbers with `get_metadata()` on one reply before scaling up. Because the codebook is identical in every request, providers that cache a shared prompt prefix will charge less for most of it; put the codebook before the survey answer, as the prompt above does.

## What to take away

- A single reply is a sample. Without `temperature` you cannot pin it down, so estimate its stability instead.
- Agreement is not accuracy. On this task three runs gave the same label for nearly every answer, and about one in six of those labels still differed from the human label.
- A codebook with definitions and decision rules does more for agreement than any prompt wording trick, and it is also where most of the remaining errors sit.
- `field_fct()` makes invalid labels impossible; it does not make wrong labels impossible.
- Repeat the call, vote, and read the rows where the runs disagree. Send those through a stronger second pass and keep a human queue for what is still unclear.
- Validate on a random, held-out, human-coded sample, and report kappa or alpha with the number of coders behind it. Treat the human labels with the same suspicion: validation found four errors in ours.
- Do not tune the codebook on the sample you validate with.

## References

- Gilardi, F., Alizadeh, M. and Kubli, M. (2023). ChatGPT outperforms crowd workers for text-annotation tasks. *PNAS* 120(30).
- Pangakis, N., Wolken, S. and Fasching, N. (2023). Automated annotation with generative AI requires validation. arXiv:2306.00176.
- Ziems, C. et al. (2024). Can large language models transform computational social science? *Computational Linguistics* 50(1).
- Barrie, C., Palaiologou, E. and Törnberg, P. (2024). Prompt stability scoring for text annotation with large language models. arXiv:2407.02039.
- Haldar, R. and Hockenmaier, J. (2025). Rating Roulette: Self-inconsistency in LLM-as-a-judge frameworks. *Findings of EMNLP 2025*; arXiv:2510.27106.
- Tam, Z. R. et al. (2024). Let me speak freely? A study on the impact of format restrictions on performance of large language models. arXiv:2408.02442.
- Krippendorff, K. (2004). *Content Analysis: An Introduction to Its Methodology*. 2nd ed. Sage.
